---
title: "Week 9 Assignment, DATA607 Fall 2016"
author: "Leland Randles"
date: "October 30, 2016"
output: html_document
---

### Web APIs

The task is to choose one of the New York Times APIs, construct an interface in R, read in some JSON data, and transform it into an R data frame.  

#### Choice of New York Times API

I chose the Movie Reviews API, for which the documentation can be found here (requires API key):  <http://developer.nytimes.com/movie_reviews_v2.json#/README>     
  
#### Construct Interface, Read JSON Data, and Transform to R Dataframe    

I installed the 'jsonlite' package to facilitate reading the JSON and transforming it to a data frame.    
  
```{r results='hide', message=FALSE, warning=FALSE}
install.packages("jsonlite",repos='http://cran.wustl.edu/')
library(jsonlite)
```  
  
The code below creates the R interface, reads in JSON data, and transforms it to a data frame.  (Code borrows heavily from this CRAN R page:  <https://cran.r-project.org/web/packages/jsonlite/vignettes/json-apis.html>)
  
```{r api, eval=TRUE}
# Use API key to create key string to be added to URL
nyt_api_key <- "&api-key=a4c5027d4f8e46e9bb938a232b600914"
# Assign API URL to variable; searches the movie reviews table using the key
# word "walk" for movies which opened between 1/1/1930 and 1/1/2015
nyt_api_url <- "http://api.nytimes.com/svc/movies/v2/reviews/search.json?query=walk&opening-date=1930-01-01;2015-01-01"
# Get JSON data from API and convert to data frame
movies <- data.frame(fromJSON(paste0(nyt_api_url, nyt_api_key)), simplifyDataFrame= TRUE)
class(movies)
# View data frame
head(movies)
```  
  
