---
title: "Week 7 Assignment, DATA607 Fall 2016"
author: "Leland Randles"
date: "October 16, 2016"
output: html_document
---

### Working With XML and JSON in R  

The task assigned is to pick three favorite books on one of our favorite subjects, and then manually create three files which contain the information:  1) an html file, 2) an xml file, and 3) a json file. Then, load those files into three separate R data frames and see if the data frames are identical.  

#### Creating Files and Posting to GitHub Pages Repository  

To facilitate the files being available to anyone on the web, I created a gh-pages branch in my GitHub repository for the course (<https://github.com/LelandoSupreme/DATA607/tree/gh-pages>) and linked that repository to a newly registered domain called <http://www.assignment7.com>. Then I committed the three files I'd created (books.html, books.xml, and books.json) to the gh-pages repository, making them available at my domain. To see the raw files, click the links below:  
  
HTML: <https://raw.githubusercontent.com/LelandoSupreme/DATA607/gh-pages/books.html>  
XML: <https://raw.githubusercontent.com/LelandoSupreme/DATA607/gh-pages/books.xml>  
JSON: <https://raw.githubusercontent.com/LelandoSupreme/DATA607/gh-pages/books.json>  
  

#### Read Files into R Data Frames    
  
To facilitate, I installed the 'XML' and 'jsonlite' packages:    
  
```{r results='hide', message=FALSE, warning=FALSE}
install.packages("XML",repos='http://cran.wustl.edu/')
library(XML)
install.packages("jsonlite",repos='http://cran.wustl.edu/')
library(jsonlite)
```  
  
First, I loaded the html file to a data frame. (The table can be viewed as displayed here: <http://assignment7.com/books.html>)  
  
```{r html, eval=TRUE}
# Use readHTMLTable to load to data frame
html_books <- data.frame(readHTMLTable("http://www.assignment7.com/books.html"))
# View result
html_books
```  
  
Second, I loaded the XML file to a data frame:  
  
```{r xml, eval=TRUE}
# Use readHTMLTable to load to data frame
xml_books <- xmlToDataFrame("http://www.assignment7.com/books.xml")
# View result
xml_books
```  
  
Lastly, I loaded the JSON file to a data frame:  
  
```{r json, eval=TRUE}
# Use readHTMLTable to load to data frame
json_books <- data.frame(fromJSON("http://www.assignment7.com/books.json", simplifyDataFrame = TRUE))
# View result
json_books
```  
  
The three data frames are identical except for the column headers. The column headers on books_html has NULL. before every column name. The books_xml column headers are just the header. The books_json column headers have books. before every column name. I am not sure why this is happening.  

Also, I doubt I am handling the authors the way the assignment intends. If I look at the sample XML file in the text (page 42), the file handles actors by using a smart tag with two attributes. I suppose the authors tab in my XML file could have included four attributes to contain up to four author names. Sames goes for the JSON file.  The value key-value pair could have had an array of authors as the value.  I am not sure how you would make a comparable HTML table, though.  